---
title: "DPR Party Vote vs Candidate Vote Analysis"
author: "Codex"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

This notebook explores the relative weight of party votes versus candidate votes in the DPR 2024 results extracted from `dpr_all.csv`. The goal is to understand which parties relied more heavily on party ballots compared to individual candidate support and how this balance varies across provinces.

# Data Preparation

```{r data-load}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(plotly)

# Utility to harmonise province names for joins
normalize_province <- function(x) {
  standardised <- x %>%
    stringr::str_to_upper() %>%
    stringr::str_replace_all("\\s+", " ") %>%
    stringr::str_trim()

  dplyr::case_when(
    standardised == "P A P U A" ~ "PAPUA",
    standardised == "KALIMANTAN TMUR" ~ "KALIMANTAN TIMUR",
    TRUE ~ standardised
  )
}

# Load the CSV generated earlier
data_path <- file.path("..", "..", "data", "processed", "dpr_all.csv")
raw <- read_csv(data_path, show_col_types = FALSE)

clean <- raw %>%
  mutate(
    PartyVote = as.numeric(PartyVote),
    CandidateVote = as.numeric(CandidateVote),
    ProvinceNorm = normalize_province(Province)
  )

unique_parties <- clean %>% distinct(PartyName) %>% arrange(PartyName)
unique_parties
```

# Party-Level Totals and Ratios

```{r party-ratio}
party_totals <- clean %>%
  distinct(ProvinceNorm, District, PartyNumber, PartyName, PartyVote) %>%
  group_by(PartyName) %>%
  summarise(PartyVoteTotal = sum(PartyVote, na.rm = TRUE), .groups = "drop")

candidate_totals <- clean %>%
  group_by(PartyName) %>%
  summarise(CandidateVoteTotal = sum(CandidateVote, na.rm = TRUE), .groups = "drop")

ratio_table <- party_totals %>%
  inner_join(candidate_totals, by = "PartyName") %>%
  mutate(Ratio = PartyVoteTotal / CandidateVoteTotal) %>%
  arrange(desc(Ratio))

ratio_table
```

# Visualising Party Reliance on Party Votes

```{r ratio-bar, fig.width=9, fig.height=5}
ratio_table %>%
  mutate(PartyName = fct_reorder(PartyName, Ratio)) %>%
  ggplot(aes(x = Ratio, y = PartyName)) +
  geom_col(fill = "#2c7fb8") +
  geom_text(aes(label = scales::percent(Ratio, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Party Vote vs Candidate Vote Ratio (National Totals)",
    x = "Party Vote / Candidate Vote",
    y = NULL,
    caption = "Ratio > 1 would indicate party votes exceed candidate votes; all parties here lean more on candidates."
  ) +
  theme_minimal(base_size = 12)
```

# Comparing Absolute Totals

```{r stacked-bar, fig.width=9, fig.height=6}
long_totals <- ratio_table %>%
  select(PartyName, PartyVoteTotal, CandidateVoteTotal) %>%
  pivot_longer(-PartyName, names_to = "VoteType", values_to = "Votes")

long_totals %>%
  mutate(
    PartyName = fct_reorder(PartyName, Votes, .fun = sum),
    VoteType = recode(VoteType,
                      PartyVoteTotal = "Party Votes",
                      CandidateVoteTotal = "Candidate Votes")
  ) %>%
  ggplot(aes(x = Votes, y = PartyName, fill = VoteType)) +
  geom_col(position = "stack") +
  scale_x_continuous(labels = scales::comma_format()) +
  scale_fill_manual(values = c("Party Votes" = "#74a9cf", "Candidate Votes" = "#0570b0")) +
  labs(
    title = "Total Votes by Party",
    x = "Votes",
    y = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 12)
```

# Province-Level Ratios

```{r province-ratios}
province_party_totals <- clean %>%
  distinct(ProvinceNorm, District, PartyNumber, PartyName, PartyVote) %>%
  group_by(ProvinceNorm) %>%
  summarise(PartyVoteTotal = sum(PartyVote, na.rm = TRUE), .groups = "drop")

province_candidate_totals <- clean %>%
  group_by(ProvinceNorm) %>%
  summarise(CandidateVoteTotal = sum(CandidateVote, na.rm = TRUE), .groups = "drop")

province_ratios <- province_party_totals %>%
  inner_join(province_candidate_totals, by = "ProvinceNorm") %>%
  mutate(Ratio = PartyVoteTotal / CandidateVoteTotal) %>%
  arrange(desc(Ratio))

province_ratios %>% head(10)
```

The table above shows the provinces with the highest reliance on party votes (top rows have the largest ratios).

# Choropleth Map of Party Vote Ratios

```{r province-map, fig.width=9, fig.height=8}
indo_sf <- ne_states(country = "Indonesia", returnclass = "sf") %>%
  select(name, geometry) %>%
  mutate(ProvinceNorm = normalize_province(name))

map_data <- indo_sf %>%
  left_join(province_ratios, by = "ProvinceNorm")

missing_provinces <- province_ratios %>%
  filter(!ProvinceNorm %in% map_data$ProvinceNorm)

if (nrow(missing_provinces) > 0) {
  message("Provinces without map geometry: ", paste(missing_provinces$ProvinceNorm, collapse = ", "))
}

map_data %>%
  ggplot() +
  geom_sf(aes(fill = Ratio), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "C", na.value = "#f0f0f0", labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Share of Party Votes Relative to Candidate Votes by Province",
    fill = "Party/Candidate",
    caption = "Grey provinces are missing from the Natural Earth boundaries (new Papua provinces)."
  ) +
  theme_void(base_size = 12) +
  theme(legend.position = "right")
```

# Party-Faceted Province Map

To compare geographic reliance for each party individually, the following faceted map shows the party-to-candidate vote ratio by province for every party. Provinces not present in the Natural Earth boundaries (e.g., newly split Papua provinces) appear in grey.

```{r party-facet-map, fig.width=12, fig.height=10}
party_province_ratios <- clean %>%
  distinct(ProvinceNorm, District, PartyName, PartyVote) %>%
  group_by(ProvinceNorm, PartyName) %>%
  summarise(PartyVoteTotal = sum(PartyVote, na.rm = TRUE), .groups = "drop") %>%
  inner_join(
    clean %>%
      group_by(ProvinceNorm, PartyName) %>%
      summarise(CandidateVoteTotal = sum(CandidateVote, na.rm = TRUE), .groups = "drop"),
    by = c("ProvinceNorm", "PartyName")
  ) %>%
  mutate(
    Ratio = if_else(CandidateVoteTotal > 0, PartyVoteTotal / CandidateVoteTotal, NA_real_)
  )

party_map_data <- indo_sf %>%
  left_join(party_province_ratios, by = "ProvinceNorm")

party_map_data %>%
  ggplot() +
  geom_sf(aes(fill = Ratio), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "C", na.value = "#f0f0f0", labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ PartyName, ncol = 3) +
  labs(
    title = "Party vs Candidate Vote Ratio by Province and Party",
    fill = "Party/Candidate",
    caption = "Grey provinces are missing in Natural Earth boundaries (e.g., newly formed Papua provinces)."
  ) +
  theme_void(base_size = 10) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
```


# Province Distribution by Party (Interactive)

```{r province-boxplot, fig.width=12, fig.height=7}
box_data <- party_province_ratios %>%
  filter(CandidateVoteTotal + PartyVoteTotal > 0) %>%
  mutate(
    Share = PartyVoteTotal / (PartyVoteTotal + CandidateVoteTotal),
    ProvinceFactor = forcats::fct_reorder(ProvinceNorm, Share, .fun = median, na.rm = TRUE),
    HoverText = paste0("Province: ", ProvinceNorm, "<br>Party: ", PartyName, "<br>Share: ", scales::percent(Share, accuracy = 0.1))
  )

box_plot <- ggplot(box_data, aes(x = ProvinceFactor, y = Share)) +
  geom_boxplot(fill = "#c7e9b4", colour = "#31a354", outlier.shape = NA, alpha = 0.6) +
  geom_point(
    aes(colour = PartyName, text = HoverText),
    position = position_jitter(width = 0.2, height = 0, seed = 123),
    size = 1.8, alpha = 0.8
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Distribution of Party Vote Share by Province",
    x = "Province",
    y = "Party Vote / Total Vote",
    colour = "Party"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotly::ggplotly(box_plot, tooltip = "text") %>%
  layout(xaxis = list(title = "Province"), yaxis = list(title = "Party Vote / Candidate Vote"))
```


# Investigating Partai Solidaritas Indonesia (PSI) Outliers

This section dives deeper into PSI's provincial party-vote dominance to determine whether the outlying shares stem from data artifacts or distinctive electoral patterns.

```{r psi-summary}
psi_province <- party_province_ratios %>%
  filter(PartyName == "Partai Solidaritas Indonesia") %>%
  mutate(Share = PartyVoteTotal / (PartyVoteTotal + CandidateVoteTotal)) %>%
  arrange(desc(Share))

psi_province %>%
  mutate(
    PartyVoteTotal = scales::comma(PartyVoteTotal),
    CandidateVoteTotal = scales::comma(CandidateVoteTotal),
    Share = scales::percent(Share, accuracy = 0.1)
  ) %>%
  select(ProvinceNorm, PartyVoteTotal, CandidateVoteTotal, Share) %>%
  head(10)
```

```{r psi-scatter, fig.width=6, fig.height=5}
psi_scatter <- party_province_ratios %>%
  filter(PartyName == "Partai Solidaritas Indonesia") %>%
  mutate(
    Share = PartyVoteTotal / (PartyVoteTotal + CandidateVoteTotal),
    HoverText = paste0("Province: ", ProvinceNorm,
                       "<br>Party votes: ", scales::comma(PartyVoteTotal),
                       "<br>Candidate votes: ", scales::comma(CandidateVoteTotal),
                       "<br>Share: ", scales::percent(Share, accuracy = 0.1))
  )

psi_scatter_plot <- ggplot(psi_scatter, aes(x = CandidateVoteTotal, y = PartyVoteTotal)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(aes(colour = Share, text = HoverText), size = 2.5) +
  scale_colour_viridis_c(option = "C", labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::comma_format()) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "PSI Party vs Candidate Votes by Province",
    x = "Candidate Votes",
    y = "Party Votes",
    colour = "Party Share"
  ) +
  theme_minimal(base_size = 11)

plotly::ggplotly(psi_scatter_plot, tooltip = "text")
```

```{r psi-boxplot-comparison, fig.width=6, fig.height=5}
comparison_parties <- party_province_ratios %>%
  filter(PartyName %in% c("Partai Solidaritas Indonesia", "Partai Buruh", "PARTAI PERINDO")) %>%
  mutate(Share = PartyVoteTotal / (PartyVoteTotal + CandidateVoteTotal))

ggplot(comparison_parties, aes(x = PartyName, y = Share, fill = PartyName)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Party Vote Share Comparison",
    x = NULL,
    y = "Party Vote Share"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

The summary table surfaces the provinces where PSI’s party ballots dominate despite modest candidate totals. The scatter plot contrasts provincial candidate and party votes, highlighting deviations from a 1:1 balance, while the boxplot juxtaposes PSI with other smaller parties (Buruh, Perindo) to show that PSI’s party-centric profile is indeed exceptional.

# Estimating DPR Winners (Heuristic)

Using the maximum candidate slate per district as a proxy for the seat allotment, the code below runs a Sainte-Laguë allocation and identifies the top personal-vote earners who would take those seats. Results should be treated as indicative only, since the actual legal seat apportionment relies on the official KPU seat counts per dapil.

```{r seat-allocation}
allocate_seats <- function(df) {
  seats <- unique(df$Seats)[1]
  if (is.na(seats) || seats <= 0) {
    return(df %>% mutate(SeatsWon = 0L))
  }

  quotients <- purrr::map2_dfr(
    df$PartyName,
    df$TotalVotes,
    ~ tibble(PartyName = .x, Quotient = .y / seq(1, by = 2, length.out = seats))
  )

  winners <- quotients %>%
    arrange(desc(Quotient)) %>%
    slice_head(n = seats) %>%
    count(PartyName, name = "SeatsWon")

  df %>%
    left_join(winners, by = "PartyName") %>%
    mutate(SeatsWon = replace_na(SeatsWon, 0L))
}

party_dapil_stats <- clean %>%
  group_by(ProvinceNorm, District, PartyName) %>%
  summarise(
    PartyBallots = first(PartyVote),
    CandidateVotes = sum(CandidateVote, na.rm = TRUE),
    TotalVotes = PartyBallots + CandidateVotes,
    CandidateCount = n(),
    .groups = "drop"
  )

seat_per_dapil <- party_dapil_stats %>%
  group_by(ProvinceNorm, District) %>%
  summarise(EstimatedSeats = max(CandidateCount, na.rm = TRUE), .groups = "drop")

seat_results <- party_dapil_stats %>%
  left_join(seat_per_dapil, by = c("ProvinceNorm", "District")) %>%
  rename(Seats = EstimatedSeats) %>%
  group_by(ProvinceNorm, District) %>%
  group_modify(~ allocate_seats(.x)) %>%
  ungroup()

dapil_totals <- seat_results %>%
  group_by(ProvinceNorm, District) %>%
  summarise(DapilTotalVotes = sum(TotalVotes), .groups = "drop")

candidate_rankings <- clean %>%
  group_by(ProvinceNorm, District, PartyName) %>%
  arrange(desc(CandidateVote), CandidateName) %>%
  mutate(ListPosition = row_number()) %>%
  ungroup()

winners <- candidate_rankings %>%
  left_join(seat_results %>% select(ProvinceNorm, District, PartyName, SeatsWon, TotalVotes),
            by = c("ProvinceNorm", "District", "PartyName")) %>%
  left_join(dapil_totals, by = c("ProvinceNorm", "District")) %>%
  mutate(SeatsWon = replace_na(SeatsWon, 0L)) %>%
  filter(ListPosition <= SeatsWon) %>%
  mutate(
    VoteShareDapil = CandidateVote / DapilTotalVotes,
    VoteShareParty = CandidateVote / TotalVotes
  ) %>%
  select(Province = ProvinceNorm, District, PartyName, CandidateName, CandidateVote, VoteShareDapil, VoteShareParty) %>%
  arrange(Province, District, PartyName, desc(CandidateVote))

national_seat_totals <- winners %>%
  count(PartyName, name = "EstimatedSeats") %>%
  arrange(desc(EstimatedSeats))

seat_outputs <- list(winners = winners, national_seat_totals = national_seat_totals)
```

```{r seat-preview}
seat_outputs$winners %>%
  mutate(
    VoteShareDapil = round(VoteShareDapil, 4),
    VoteShareParty = round(VoteShareParty, 4)
  ) %>%
  head(20)
```

```{r seat-summary}
seat_outputs$national_seat_totals
```

The preview above lists the first 20 projected MPs, while the summary tallies estimated seats per party. Consider replacing the heuristic seat totals with the official dapil seat counts when they become available for precise certification.


# Winners Vote Share Scatter

```{r winners-scatter, fig.width=10, fig.height=6}
winner_scatter <- seat_outputs$winners

scatter_plot <- ggplot(winner_scatter, aes(x = VoteShareParty, y = VoteShareDapil,colour = Party)) +
  geom_jitter(aes(text = paste0(
    "Province: ", Province,
    "
District: ", District,
    "
Party: ", PartyName,
    "
Candidate: ", CandidateName,
    "
Vote share (dapil): ", scales::percent(VoteShareDapil, accuracy = 0.1),
    "
Vote share (party): ", scales::percent(VoteShareParty, accuracy = 0.1)
  )),
  width = 0.2, alpha = 0.6, size = 1.8, colour = "#2b8cbe") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Winning Candidates' Share of Total Valid Votes per Dapil",
    x = "Party",
    y = "Vote Share (Dapil)"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotly::ggplotly(scatter_plot, tooltip = "text")
```

# Dapil Representation Coverage

```{r dapil-coverage, fig.width=11, fig.height=6}
dapil_coverage <- seat_outputs$winners %>%
  group_by(Province, District) %>%
  summarise(
    TotalVoteShare = sum(VoteShareDapil, na.rm = TRUE),
    Seats = n(),
    Parties = n_distinct(PartyName),
    .groups = "drop"
  ) %>%
  mutate(
    DapilLabel = paste(Province, District, sep = " - ")
  )

coverage_plot <- ggplot(dapil_coverage, aes(x = reorder(DapilLabel, TotalVoteShare), y = TotalVoteShare)) +
  geom_col(fill = "#74add1") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Winners' Combined Vote Share by Dapil",
    x = "Dapil",
    y = "Sum of Winners' Vote Share (Dapil)",
    caption = "Sum of elected candidates' vote share – higher values indicate a larger portion of voters are represented by winning candidates"
  ) +
  theme_minimal(base_size = 11)

plotly::ggplotly(coverage_plot, tooltip = c("x", "y"))
```

# Party Vote Coverage by Dapil

```{r dapil-party-coverage, fig.width=11, fig.height=6}
party_coverage <- seat_results %>%
  left_join(dapil_totals, by = c("ProvinceNorm", "District")) %>%
  filter(SeatsWon > 0) %>%
  group_by(ProvinceNorm, District) %>%
  summarise(
    PartyVoteShare = sum(TotalVotes, na.rm = TRUE) / first(DapilTotalVotes),
    SeatsWon = sum(SeatsWon, na.rm = TRUE),
    PartiesRepresented = n(),
    .groups = "drop"
  ) %>%
  mutate(DapilLabel = paste(ProvinceNorm, District, sep = " - "))

party_plot <- ggplot(party_coverage, aes(x = reorder(DapilLabel, PartyVoteShare), y = PartyVoteShare)) +
  geom_col(fill = "#4575b4") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Share of Valid Votes Captured by Parties Winning Seats",
    x = "Dapil",
    y = "Party Vote Share (Total Votes)",
    caption = "Calculated from heuristic seat allocation; reflects combined party ballots plus candidate votes for parties securing seats"
  ) +
  theme_minimal(base_size = 11)

plotly::ggplotly(party_plot, tooltip = c("x", "y"))
```

# Discussion

```{r summary-table}
ratio_table %>%
  mutate(
    PartyVoteTotal = scales::comma(PartyVoteTotal),
    CandidateVoteTotal = scales::comma(CandidateVoteTotal),
    Ratio = scales::number(Ratio, accuracy = 0.0001)
  ) %>%
  arrange(desc(as.numeric(gsub(",", "", CandidateVoteTotal))))
```

Key observations:

- Parties such as Partai Buruh, Partai Gelombang Rakyat Indonesia, and PSI have the highest ratios, indicating stronger reliance on party-brand votes compared to their candidate ballots.
- Major vote-getters (e.g., PDI-P, Golkar, Gerindra) gather large absolute totals but with lower ratios, signalling that their success is driven primarily by candidate vote counts.
- No party crosses a ratio of 1, underscoring that across the board candidate votes outweigh party votes in absolute terms.
- Provincially, areas with newly formed provinces (e.g., Papua Tengah, Papua Pegunungan) are not yet represented in the Natural Earth shapefile and therefore appear as grey gaps on the map.

# Next Steps

Further analyses could look at provincial or district-level variance by computing ratios per region, or correlate candidate vote totals with seat outcomes. The province ratios calculated above can also be broken down by party to study geographic strongholds.
