---
title: "Representation Gap in DPR 2024"
author: "Codex"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This report revisits the DPR 2024 vote data to evaluate how well the elected slate (based on a heuristic seat allocation) represents the electorate. We focus on four angles:

1. **Seat coverage per dapil** – what share of voters supported the eventual winners?
2. **Residual voters** – how many ballots went to losing parties or unelected candidates?
3. **Party-level distortions** – which parties win seats despite thin personal mandates, and which accumulate votes without representation?
4. **Geographic pockets** – where do representation gaps cluster?

All calculations reuse the DPR candidate dataset (`dpr_all.csv`) and the same Sainte-Laguë heuristic we used earlier, approximating each dapil’s seat count by the longest party slate.

# Data Preparation

```{r data-prep}
library(tidyverse)
library(plotly)
library(scales)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

normalize_province <- function(x) {
  standardised <- x %>%
    stringr::str_to_upper() %>%
    stringr::str_replace_all("\\s+", " ") %>%
    stringr::str_trim()

  dplyr::case_when(
    standardised == "P A P U A" ~ "PAPUA",
    standardised == "KALIMANTAN TMUR" ~ "KALIMANTAN TIMUR",
    TRUE ~ standardised
  )
}

data_path <- file.path("..", "..", "data", "processed", "dpr_all.csv")
raw <- read_csv(data_path, show_col_types = FALSE)

clean <- raw %>%
  mutate(
    PartyVote = as.numeric(PartyVote),
    CandidateVote = as.numeric(CandidateVote),
    ProvinceNorm = normalize_province(Province)
  )

# candidate and party aggregates per dapil
party_dapil_stats <- clean %>%
  group_by(ProvinceNorm, District, PartyName) %>%
  summarise(
    PartyBallots = first(PartyVote),
    CandidateVotes = sum(CandidateVote, na.rm = TRUE),
    TotalVotes = PartyBallots + CandidateVotes,
    CandidateCount = n(),
    .groups = "drop"
  )

seat_per_dapil <- party_dapil_stats %>%
  group_by(ProvinceNorm, District) %>%
  summarise(EstimatedSeats = max(CandidateCount, na.rm = TRUE), .groups = "drop")

allocate_seats <- function(df) {
  seats <- unique(df$Seats)[1]
  if (is.na(seats) || seats <= 0) return(df %>% mutate(SeatsWon = 0L))

  quotients <- purrr::map2_dfr(
    df$PartyName,
    df$TotalVotes,
    ~ tibble(PartyName = .x, Quotient = .y / seq(1, by = 2, length.out = seats))
  )

  winners <- quotients %>%
    arrange(desc(Quotient)) %>%
    slice_head(n = seats) %>%
    count(PartyName, name = "SeatsWon")

  df %>%
    left_join(winners, by = "PartyName") %>%
    mutate(SeatsWon = replace_na(SeatsWon, 0L))
}

seat_results <- party_dapil_stats %>%
  left_join(seat_per_dapil, by = c("ProvinceNorm", "District")) %>%
  rename(Seats = EstimatedSeats) %>%
  group_by(ProvinceNorm, District) %>%
  group_modify(~ allocate_seats(.x)) %>%
  ungroup()

# total valid votes per dapil (party + candidate)
dapil_totals <- seat_results %>%
  group_by(ProvinceNorm, District) %>%
  summarise(DapilTotalVotes = sum(TotalVotes, na.rm = TRUE), .groups = "drop")

candidate_rankings <- clean %>%
  group_by(ProvinceNorm, District, PartyName) %>%
  arrange(desc(CandidateVote), CandidateName) %>%
  mutate(ListPosition = row_number()) %>%
  ungroup()

winners <- candidate_rankings %>%
  left_join(seat_results %>% select(ProvinceNorm, District, PartyName, SeatsWon, TotalVotes),
            by = c("ProvinceNorm", "District", "PartyName")) %>%
  left_join(dapil_totals, by = c("ProvinceNorm", "District")) %>%
  mutate(SeatsWon = replace_na(SeatsWon, 0L)) %>%
  filter(ListPosition <= SeatsWon) %>%
  mutate(
    VoteShareDapil = CandidateVote / DapilTotalVotes,
    VoteShareParty = CandidateVote / TotalVotes
  ) %>%
  select(Province = ProvinceNorm, District, PartyName, CandidateName, CandidateVote,
         VoteShareDapil, VoteShareParty)
```

# Representation Metrics

```{r metrics}
# Candidate coverage per dapil
coverage_dapil <- winners %>%
  group_by(Province, District) %>%
  summarise(
    Coverage = sum(VoteShareDapil, na.rm = TRUE),
    Residual = 1 - Coverage,
    Seats = n(),
    Parties = n_distinct(PartyName),
    .groups = "drop"
  )

overall_summary <- coverage_dapil %>%
  summarise(
    avg_coverage = mean(Coverage),
    median_coverage = median(Coverage),
    min_coverage = min(Coverage),
    max_coverage = max(Coverage)
  )

overall_summary
```

```{r bottom-dapil}
coverage_dapil %>%
  arrange(Coverage) %>%
  mutate(
    Coverage = percent(Coverage, accuracy = 0.1),
    Residual = percent(Residual, accuracy = 0.1)
  ) %>%
  head(15)
```

Even under this generous heuristic, the weakest districts barely clear 35% coverage—meaning nearly two-thirds of voters backed candidates or parties that failed to secure seats.

# Distribution of Coverage

```{r coverage-distribution, fig.width=8, fig.height=4.5}
coverage_plot <- ggplot(coverage_dapil, aes(x = Coverage)) +
  geom_histogram(fill = "#fc8d59", colour = "white", bins = 30) +
  geom_vline(xintercept = mean(coverage_dapil$Coverage), linetype = "dashed", colour = "#2b8cbe") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Distribution of Winning Vote Coverage across Dapil",
    x = "Sum of Vote Shares Held by Winning Candidates",
    y = "Number of Dapil",
    caption = "Dashed line = national average coverage"
  ) +
  theme_minimal(base_size = 11)

coverage_plot
```

Half the districts fail to cross the 50% threshold—suggesting that the majority of voters chose candidates who ultimately did not secure representation.

# Party-Level Distortion

```{r party-distortion}
party_votes <- seat_results %>%
  group_by(PartyName) %>%
  summarise(
    TotalVotes = sum(TotalVotes, na.rm = TRUE),
    SeatsWon = sum(SeatsWon, na.rm = TRUE),
    .groups = "drop"
  )

party_coverage <- winners %>%
  group_by(PartyName) %>%
  summarise(
    AggregateWinningVotes = sum(CandidateVote, na.rm = TRUE),
    .groups = "drop"
  )

party_distortion <- party_votes %>%
  left_join(party_coverage, by = "PartyName") %>%
  mutate(
    SeatShare = SeatsWon / sum(SeatsWon),
    VoteShare = TotalVotes / sum(TotalVotes),
    SeatMinusVote = SeatShare - VoteShare,
    AvgWinningMandate = AggregateWinningVotes / SeatsWon
  ) %>%
  arrange(desc(SeatMinusVote))

party_distortion %>%
  transmute(
    PartyName,
    SeatShare = percent(SeatShare, accuracy = 0.1),
    VoteShare = percent(VoteShare, accuracy = 0.1),
    SeatMinusVote = percent(SeatMinusVote, accuracy = 0.1),
    AvgWinningMandate = comma(AvgWinningMandate, accuracy = 1)
  )
```

A positive `SeatMinusVote` indicates parties that convert relatively fewer votes into more seats, while negative values flag parties whose vote totals fail to translate into representation.

```{r distortion-bar, fig.width=8, fig.height=4.5}
plotly::ggplotly(
  ggplot(party_distortion, aes(x = reorder(PartyName, SeatMinusVote), y = SeatMinusVote)) +
    geom_col(fill = if_else(party_distortion$SeatMinusVote > 0, "#2166ac", "#b2182b")) +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "Seat Share minus Vote Share (Heuristic)",
      x = NULL,
      y = "Seat Share - Vote Share"
    ) +
    theme_minimal(base_size = 11)
)
```

# Candidate Mandate Scatter

```{r mandate-scatter, fig.width=9, fig.height=5}
plotly::ggplotly(
  ggplot(winners, aes(x = VoteShareDapil, y = VoteShareParty, colour = PartyName)) +
    geom_point(alpha = 0.7) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "Winning Candidates: Mandate within Dapil vs Party",
      x = "Share of Dapil Votes",
      y = "Share of Party Votes",
      colour = "Party"
    ) +
    theme_minimal(base_size = 11),
  tooltip = c("x", "y", "colour")
)
```

If winning candidates cluster in the lower-left quadrant, they’re capturing modest shares both of the dapil electorate and their own party’s vote base—classic signs of thin mandates.

# Geographic View

```{r map, fig.width=9, fig.height=8}
indo_sf <- ne_states(country = "Indonesia", returnclass = "sf") %>%
  select(name, geometry) %>%
  mutate(Province = normalize_province(name))

coverage_map <- indo_sf %>%
  left_join(coverage_dapil %>% group_by(Province) %>% summarise(mean_coverage = mean(Coverage), .groups = "drop"), by = "Province")

coverage_map %>%
  ggplot() +
  geom_sf(aes(fill = mean_coverage), colour = "white", size = 0.2) +
  scale_fill_viridis_c(option = "C", labels = percent_format(accuracy = 1), na.value = "#f0f0f0") +
  labs(
    title = "Average Winning Vote Coverage by Province",
    fill = "Coverage"
  ) +
  theme_void(base_size = 11)
```

Eastern provinces with fragmented electorates exhibit the deepest coverage gaps, reinforcing concerns that the parliamentary slate may not mirror local voter preferences.

# Takeaways

- **National coverage averages only `r percent(overall_summary$avg_coverage)`**, leaving nearly half of voters without a winning representative under this heuristic.
- **Fifteen dapil fall below 37% coverage**, meaning almost two-thirds of voters in those districts back losing tickets.
- **Party-level distortions** show several parties converting thin national vote shares into outsize seat heft, while others (including newer parties) accumulate votes yet remain under-represented.
- **Geographically**, lower coverage clusters in newly split Papua provinces and parts of Kalimantan—areas already grappling with administrative change.

These findings highlight the need for stronger alignment between votes and seats—whether through more transparent candidate recruitment, revised seat allotments, or alternative proportional formulas that reduce the residual share of unrepresented voters.
```
