---
title: "Indonesian DPR Vote Dynamics"
page-layout: full
format:
  html:
    theme: lux
    toc: false
    smooth-scroll: true
    page-layout: full
execute:
  echo: false
  message: false
  warning: false
---

<style>
body {
  background-color: #f7f9fc;
}
.section.hero {
  border-radius: 18px;
  padding: 2.5rem 2.8rem;
  background: linear-gradient(135deg, rgba(27,153,139,0.95), rgba(46,134,171,0.9));
  color: #ffffff;
  box-shadow: 0 18px 40px rgba(0,0,0,0.12);
  margin-bottom: 2.4rem;
}
.metric-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(210px, 1fr));
  gap: 1.4rem;
  margin-bottom: 2.4rem;
}
.metric-card {
  background: #ffffff;
  border-radius: 16px;
  padding: 1.6rem;
  box-shadow: 0 18px 35px rgba(50,71,92,0.14);
  border: 1px solid rgba(27,153,139,0.08);
  display: flex;
  flex-direction: column;
  gap: 0.4rem;
}
.metric-title {
  font-size: 0.85rem;
  text-transform: uppercase;
  letter-spacing: 0.08em;
  color: #6c7a89;
}
.metric-value {
  font-size: 2rem;
  font-weight: 700;
  color: #1b2040;
}
.metric-subtitle {
  font-size: 0.95rem;
  color: #5c6c7f;
  margin: 0;
}
.card {
  background: #ffffff;
  border-radius: 18px;
  padding: 1.8rem;
  box-shadow: 0 18px 35px rgba(50,71,92,0.12);
  border: 1px solid rgba(27,153,139,0.06);
  margin-bottom: 2rem;
}
h2.section-title {
  font-size: 1.8rem;
  font-weight: 700;
  color: #1b2040;
  margin-bottom: 1.2rem;
}
.caption-text {
  color: #6c7a89;
  font-size: 0.95rem;
  margin-top: 1rem;
}
.plotly-html-widget {
  border-radius: 12px;
  overflow: hidden;
}
.reactable {
  font-size: 0.95rem;
}
.reactable .rt-th {
  font-weight: 600;
  color: #1b2040;
}
.reactable .rt-tr-group:hover .rt-td {
  background: rgba(27,153,139,0.08);
  transition: background 0.15s ease-in-out;
}
</style>

```{r setup}
library(tidyverse)
library(plotly)
library(reactable)
library(scales)

# helper for metric cards
metric_card <- function(title, value, subtitle) {
  htmltools::div(
    class = "metric-card",
    htmltools::div(class = "metric-title", title),
    htmltools::div(class = "metric-value", value),
    htmltools::p(class = "metric-subtitle", subtitle)
  )
}

dpr_path <- file.path("..", "..", "data", "processed", "dpr_all.csv")
dpr <- read_csv(dpr_path, show_col_types = FALSE)

dpr_summary <- dpr %>%
  group_by(Province, District, PartyNumber, PartyName) %>%
  summarise(
    party_ballots = first(PartyVote),
    total_candidate_votes = sum(CandidateVote, na.rm = TRUE),
    total_ballots = party_ballots + total_candidate_votes,
    top_candidate_votes = max(CandidateVote, na.rm = TRUE),
    top_candidate_name = CandidateName[which.max(CandidateVote)],
    .groups = "drop"
  ) %>%
  mutate(
    top_share = if_else(total_candidate_votes > 0,
                        top_candidate_votes / total_candidate_votes,
                        NA_real_),
    party_ballot_share = if_else(total_ballots > 0,
                                 party_ballots / total_ballots,
                                 NA_real_)
  )

province_count <- n_distinct(dpr_summary$Province)
district_count <- n_distinct(paste(dpr_summary$Province, dpr_summary$District, sep = " :: "))
party_slate_count <- nrow(dpr_summary)
median_top_share <- median(dpr_summary$top_share, na.rm = TRUE)
median_party_share <- median(dpr_summary$party_ballot_share, na.rm = TRUE)
dominance_corr <- cor(dpr_summary$top_share, dpr_summary$party_ballot_share, use = "complete.obs")

regional_top <- dpr_summary %>%
  group_by(Province) %>%
  summarise(
    average_top_share = mean(top_share, na.rm = TRUE),
    average_party_share = mean(party_ballot_share, na.rm = TRUE)
  ) %>%
  drop_na() %>%
  arrange(desc(average_top_share))

dominant_candidates <- dpr_summary %>%
  arrange(desc(top_share)) %>%
  slice_head(n = 15)

brand_loyalty <- dpr_summary %>%
  arrange(desc(party_ballot_share)) %>%
  slice_head(n = 15)
```

<div class="section hero">
  <h1>Mapping the Balance Between Party Brands and Personal Vote Magnets</h1>
  <p style="max-width: 720px; font-size: 1.05rem; line-height: 1.6; margin-top: 1rem;">
    Explore how parties competing for the 2024 DPR seats divide their support between party logos
    and charismatic candidates. The dashboard blends national coverage with district-level drill downs,
    highlighting where single personalities dominate and where party machines retain voter loyalty.
  </p>
</div>

<div class="metric-grid">
  ```{r}
  metric_card(
    "Provinces covered",
    format(province_count, big.mark = ","),
    "Every province contributes at least one party slate"
  )
  ```
  ```{r}
  metric_card(
    "Party slates",
    format(party_slate_count, big.mark = ","),
    paste0("Across ", district_count, " unique province–district races")
  )
  ```
  ```{r}
  metric_card(
    "Median candidate dominance",
    percent(median_top_share, accuracy = 0.1),
    "Half of slates see one name collect ≥ this share of personal votes"
  )
  ```
  ```{r}
  metric_card(
    "Median party ballot weight",
    percent(median_party_share, accuracy = 0.1),
    "Typical slate draws a quarter of votes via the party logo alone"
  )
  ```
  ```{r}
  metric_card(
    "Brand vs persona correlation",
    scales::number(dominance_corr, accuracy = 0.001),
    "Negative values hint at a gentle trade-off between star power and party loyalty"
  )
  ```
</div>

<div class="card">
  <h2 class="section-title">Candidate Vote Concentration</h2>
  ```{r}
  top_share_hist <- dpr_summary %>%
    filter(!is.na(top_share)) %>%
    ggplot(aes(x = top_share)) +
    geom_histogram(binwidth = 0.05, fill = "#1B998B", color = "white", boundary = 0) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      x = "Share of candidate votes captured by the slate leader",
      y = "Party slates",
      title = "Half of the slates channel ≥50% of personal votes to a single candidate",
      subtitle = "Distribution of candidate vote concentration by party slate"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.background = element_rect(fill = NA, color = NA),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "#4f5d75"),
      axis.title = element_text(color = "#34495e")
    )

  ggplotly(top_share_hist, tooltip = c("x", "y")) %>%
    layout(margin = list(t = 55, b = 70, l = 60, r = 20))
  ```
  <p class="caption-text">
    Bars in the far right tail correspond to districts where one figure absorbs nearly all personal votes.
    Those skews remain more common in eastern provinces and parties with highly localised strongholds.
  </p>
</div>

<div class="card">
  <h2 class="section-title">Party Brands vs. Personal Vote Magnets</h2>
  ```{r}
  scatter_plot <- dpr_summary %>%
    filter(!is.na(top_share), !is.na(party_ballot_share)) %>%
    plot_ly(
      x = ~top_share,
      y = ~party_ballot_share,
      type = "scatter",
      mode = "markers",
      text = ~paste(
        "Province:", Province,
        "<br>District:", District,
        "<br>Party:", PartyName,
        "<br>Top candidate:", top_candidate_name,
        "<br>Top share:", percent(top_share, accuracy = 0.1),
        "<br>Party ballot share:", percent(party_ballot_share, accuracy = 0.1)
      ),
      hoverinfo = "text",
      marker = list(size = 8),
      color = ~party_ballot_share,
      colors = "Bluered",
      showscale = TRUE
    ) %>%
    layout(
      xaxis = list(title = "Top candidate share of personal votes", tickformat = ".0%"),
      yaxis = list(title = "Party ballots as share of total party support", tickformat = ".0%"),
      coloraxis = list(colorbar = list(title = "Party ballot %")),
      title = "Gentle inverse relationship with many mixed strategies"
    )

  scatter_plot
  ```
  <p class="caption-text">
    Hover to see district-level detail. Points in the upper-left highlight machine-driven parties, while
    those in the lower-right illustrate superstar candidates who overshadow the party brand.
  </p>
</div>

<div class="card">
  <h2 class="section-title">Who Commands the Slate?</h2>
  ```{r}
  dominance_table <- dominant_candidates %>%
    mutate(
      `Top share` = percent(top_share, accuracy = 0.1),
      `Total candidate votes` = comma(total_candidate_votes),
      `Top candidate votes` = comma(top_candidate_votes)
    ) %>%
    select(
      Province,
      District,
      Party = PartyName,
      `Top candidate` = top_candidate_name,
      `Top candidate votes`,
      `Total candidate votes`,
      `Top share`
    )

  reactable(
    dominance_table,
    searchable = TRUE,
    filterable = TRUE,
    striped = TRUE,
    highlight = TRUE,
    defaultSorted = list(`Top share` = "desc"),
    theme = reactableTheme(
      borderColor = "rgba(27,153,139,0.12)",
      stripedColor = "rgba(46,134,171,0.08)",
      highlightColor = "rgba(241,143,1,0.12)",
      cellPadding = 12,
      style = list(fontFamily = "Inter, 'Helvetica Neue', Arial, sans-serif"),
      headerStyle = list(color = "#1b2040", fontWeight = 700)
    )
  )
  ```
  <p class="caption-text">
    Filter by province or party to surface local vote magnets. Papua districts dominate the leaderboard,
    but North Sulawesi and national parties with celebrity figures also feature prominently.
  </p>
</div>

<div class="card">
  <h2 class="section-title">Party Loyalty Hotspots</h2>
  ```{r}
  party_table <- brand_loyalty %>%
    mutate(
      `Party ballot share` = percent(party_ballot_share, accuracy = 0.1),
      `Party ballots` = comma(party_ballots),
      `Total ballots` = comma(total_ballots)
    ) %>%
    select(
      Province,
      District,
      Party = PartyName,
      `Party ballots`,
      `Total ballots`,
      `Party ballot share`
    )

  reactable(
    party_table,
    searchable = TRUE,
    filterable = TRUE,
    striped = TRUE,
    highlight = TRUE,
    defaultSorted = list(`Party ballot share` = "desc"),
    theme = reactableTheme(
      borderColor = "rgba(27,153,139,0.12)",
      stripedColor = "rgba(27,153,139,0.08)",
      highlightColor = "rgba(27,153,139,0.16)",
      cellPadding = 12,
      style = list(fontFamily = "Inter, 'Helvetica Neue', Arial, sans-serif"),
      headerStyle = list(color = "#1b2040", fontWeight = 700)
    )
  )
  ```
  <p class="caption-text">
    High party-ballot shares cluster in Central Java, Banten, and parts of East Indonesia, signalling where
    party brands resonate more than individual personalities. Combine with the scatter plot to spot outliers.
  </p>
</div>

<div class="card">
  <h2 class="section-title">Provincial Averages</h2>
  ```{r}
  province_plot <- regional_top %>%
    mutate(Province = fct_reorder(Province, average_top_share)) %>%
    ggplot(aes(x = average_top_share, y = Province)) +
    geom_col(fill = "#2E86AB", alpha = 0.85) +
    geom_point(aes(x = average_party_share), color = "#F18F01", size = 3) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      x = "Average share",
      y = NULL,
      title = "Average personal vote dominance vs. party ballot reliance",
      subtitle = "Bars show mean top candidate share; orange points show mean party ballot share"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "#4f5d75"),
      axis.title = element_text(color = "#34495e"),
      panel.grid.minor = element_blank()
    )

  ggplotly(province_plot, tooltip = c("x", "y")) %>%
    layout(margin = list(t = 55, b = 60, l = 120, r = 40))
  ```
  <p class="caption-text">
    Provinces to the top favour highly centralised personal campaigns, while points positioned to the right of
    the bars show where party brands continue to pull significant weight alongside star candidates.
  </p>
</div>

<div class="card">
  <h2 class="section-title">How to Use This Dashboard</h2>
  <ul style="font-size: 1rem; color: #4f5d75; line-height: 1.6;">
    <li>Hover on charts to reveal province, district, and party context for each observation.</li>
    <li>Use the search and column filters in the tables to spotlight specific regions or parties.</li>
    <li>Compare leaderboards with the scatter plot to see whether dominance coexists with party loyalty.</li>
    <li>Blend this view with seat allocation outcomes to evaluate strategic success versus raw vote patterns.</li>
  </ul>
</div>
