---
title: "Estimated DPR Winners by Party"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
library(tidyverse)
library(scales)

data_path <- file.path("..", "..", "data", "processed", "dpr_all.csv")
raw <- read_csv(data_path, show_col_types = FALSE)

normalize_province <- function(x) {
  standardised <- x %>%
    stringr::str_to_upper() %>%
    stringr::str_replace_all("\\s+", " ") %>%
    stringr::str_trim()

  case_when(
    standardised == "P A P U A" ~ "PAPUA",
    standardised == "KALIMANTAN TMUR" ~ "KALIMANTAN TIMUR",
    TRUE ~ standardised
  )
}

clean <- raw %>%
  mutate(
    PartyVote = as.numeric(PartyVote),
    CandidateVote = as.numeric(CandidateVote),
    ProvinceNorm = normalize_province(Province)
  )

party_dapil_stats <- clean %>%
  group_by(ProvinceNorm, District, PartyName) %>%
  summarise(
    PartyBallots = first(PartyVote),
    CandidateVotes = sum(CandidateVote, na.rm = TRUE),
    TotalVotes = PartyBallots + CandidateVotes,
    CandidateCount = n(),
    .groups = "drop"
  )

seat_per_dapil <- party_dapil_stats %>%
  group_by(ProvinceNorm, District) %>%
  summarise(EstimatedSeats = max(CandidateCount, na.rm = TRUE), .groups = "drop")

allocate_seats <- function(df) {
  seats <- unique(df$Seats)[1]
  if (is.na(seats) || seats <= 0) {
    return(df %>% mutate(SeatsWon = 0L))
  }

  quotients <- map2_dfr(
    df$PartyName,
    df$TotalVotes,
    ~ tibble(PartyName = .x, Quotient = .y / seq(1, by = 2, length.out = seats))
  )

  winners <- quotients %>%
    arrange(desc(Quotient)) %>%
    slice_head(n = seats) %>%
    count(PartyName, name = "SeatsWon")

  df %>%
    left_join(winners, by = "PartyName") %>%
    mutate(SeatsWon = replace_na(SeatsWon, 0L))
}

seat_results <- party_dapil_stats %>%
  left_join(seat_per_dapil, by = c("ProvinceNorm", "District")) %>%
  rename(Seats = EstimatedSeats) %>%
  group_by(ProvinceNorm, District) %>%
  group_modify(~ allocate_seats(.x)) %>%
  ungroup()

candidate_rankings <- clean %>%
  group_by(ProvinceNorm, District, PartyName) %>%
  arrange(desc(CandidateVote), CandidateName) %>%
  mutate(ListPosition = row_number()) %>%
  ungroup()

winners <- candidate_rankings %>%
  left_join(seat_results %>% select(ProvinceNorm, District, PartyName, SeatsWon, TotalVotes),
            by = c("ProvinceNorm", "District", "PartyName")) %>%
  mutate(SeatsWon = replace_na(SeatsWon, 0L)) %>%
  filter(ListPosition <= SeatsWon, SeatsWon > 0) %>%
  mutate(
    VoteShareParty = if_else(TotalVotes > 0, CandidateVote / TotalVotes, NA_real_),
    Province = ProvinceNorm
  ) %>%
  select(Province, District, PartyName, CandidateName, CandidateVote, VoteShareParty)

winners_by_party <- seat_results %>%
  group_by(PartyName) %>%
  summarise(EstimatedSeats = sum(SeatsWon, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(EstimatedSeats))

total_seats_original <- sum(winners_by_party$EstimatedSeats, na.rm = TRUE)

coalition_threshold <- 0.04
coalition_label <- "Non-Parliamentary Coalition"

party_vote_share <- party_dapil_stats %>%
  group_by(PartyName) %>%
  summarise(TotalVotesNational = sum(TotalVotes, na.rm = TRUE), .groups = "drop") %>%
  mutate(Share = TotalVotesNational / sum(TotalVotesNational, na.rm = TRUE))

coalition_parties <- party_vote_share %>%
  filter(Share < coalition_threshold) %>%
  arrange(desc(Share)) %>%
  pull(PartyName)

party_dapil_coalition <- party_dapil_stats %>%
  mutate(PartyName = if_else(PartyName %in% coalition_parties, coalition_label, PartyName)) %>%
  group_by(ProvinceNorm, District, PartyName) %>%
  summarise(
    PartyBallots = sum(PartyBallots, na.rm = TRUE),
    CandidateVotes = sum(CandidateVotes, na.rm = TRUE),
    TotalVotes = sum(TotalVotes, na.rm = TRUE),
    CandidateCount = sum(CandidateCount, na.rm = TRUE),
    .groups = "drop"
  )

seat_results_coalition <- party_dapil_coalition %>%
  left_join(seat_per_dapil, by = c("ProvinceNorm", "District")) %>%
  rename(Seats = EstimatedSeats) %>%
  group_by(ProvinceNorm, District) %>%
  group_modify(~ allocate_seats(.x)) %>%
  ungroup()

coalition_seats_by_dapil <- seat_results_coalition %>%
  filter(PartyName == coalition_label) %>%
  transmute(ProvinceNorm, District, CoalitionSeats = SeatsWon)

coalition_internal_seat_results <- party_dapil_stats %>%
  filter(PartyName %in% coalition_parties) %>%
  left_join(coalition_seats_by_dapil, by = c("ProvinceNorm", "District")) %>%
  filter(!is.na(CoalitionSeats), CoalitionSeats > 0) %>%
  # re-run Sainte-Laguë within each dapil using only coalition parties
  mutate(Seats = CoalitionSeats) %>%
  group_by(ProvinceNorm, District) %>%
  group_modify(~ allocate_seats(.x)) %>%
  ungroup() %>%
  select(-Seats)

non_coalition_winners <- candidate_rankings %>%
  filter(!(PartyName %in% coalition_parties)) %>%
  left_join(
    seat_results_coalition %>%
      filter(PartyName != coalition_label) %>%
      select(ProvinceNorm, District, PartyName, SeatsWon, TotalVotes),
    by = c("ProvinceNorm", "District", "PartyName")
  ) %>%
  mutate(SeatsWon = replace_na(SeatsWon, 0L)) %>%
  filter(ListPosition <= SeatsWon, SeatsWon > 0) %>%
  mutate(
    CoalitionParty = PartyName,
    VoteShare = if_else(TotalVotes > 0, CandidateVote / TotalVotes, NA_real_),
    Province = ProvinceNorm,
    OriginalParty = PartyName
  ) %>%
  select(Province, District, CoalitionParty, OriginalParty, CandidateName, CandidateVote, VoteShare)

coalition_winners <- candidate_rankings %>%
  filter(PartyName %in% coalition_parties) %>%
  left_join(
    coalition_internal_seat_results %>%
      select(ProvinceNorm, District, PartyName, SeatsWon, TotalVotes),
    by = c("ProvinceNorm", "District", "PartyName")
  ) %>%
  mutate(SeatsWon = replace_na(SeatsWon, 0L)) %>%
  filter(ListPosition <= SeatsWon, SeatsWon > 0) %>%
  mutate(
    CoalitionParty = coalition_label,
    VoteShare = if_else(TotalVotes > 0, CandidateVote / TotalVotes, NA_real_),
    Province = ProvinceNorm,
    OriginalParty = PartyName
  ) %>%
  select(Province, District, CoalitionParty, OriginalParty, CandidateName, CandidateVote, VoteShare)

winners_coalition <- bind_rows(non_coalition_winners, coalition_winners)

winners_by_party_coalition <- seat_results_coalition %>%
  group_by(PartyName) %>%
  summarise(EstimatedSeats = sum(SeatsWon, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(EstimatedSeats))

total_seats_coalition <- sum(winners_by_party_coalition$EstimatedSeats, na.rm = TRUE)

coalition_breakdown <- coalition_internal_seat_results %>%
  group_by(OriginalParty = PartyName) %>%
  summarise(EstimatedSeats = sum(SeatsWon, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(EstimatedSeats))

presidential_map <- c(
  "Partai Kebangkitan Bangsa" = "Anies",
  "Partai Keadilan Sejahtera" = "Anies",
  "Partai NasDem" = "Anies",
  "Partai Ummat" = "Anies",
  "Partai Golongan Karya" = "Prabowo",
  "Partai Gerakan Indonesia Raya" = "Prabowo",
  "Partai Amanat Nasional" = "Prabowo",
  "Partai Bulan Bintang" = "Prabowo",
  "Partai Demokrat" = "Prabowo",
  "Partai Gelombang Rakyat Indonesia" = "Prabowo",
  "Partai Garda Republik Indonesia" = "Prabowo",
  "Partai Solidaritas Indonesia" = "Prabowo",
  "Partai Demokrasi Indonesia Perjuangan" = "Ganjar",
  "Partai Hati Nurani Rakyat" = "Ganjar",
  "Partai Persatuan Pembangunan" = "Ganjar",
  "PARTAI PERINDO" = "Ganjar",
  "Partai Buruh" = "Independent",
  "Partai Kebangkitan Nusantara" = "Independent"
)

add_presidential_coalition <- function(df, party_col = "PartyName", new_col = "PresidentialBloc") {
  df %>% mutate({{ new_col }} := presidential_map[.data[[party_col]]])
}

presidential_dapil_stats <- party_dapil_stats %>%
  add_presidential_coalition(party_col = "PartyName", new_col = "PresidentialBloc") %>%
  mutate(PresidentialBloc = replace_na(PresidentialBloc, "Other")) %>%
  group_by(ProvinceNorm, District, PresidentialBloc) %>%
  summarise(
    PartyBallots = sum(PartyBallots, na.rm = TRUE),
    CandidateVotes = sum(CandidateVotes, na.rm = TRUE),
    TotalVotes = sum(TotalVotes, na.rm = TRUE),
    CandidateCount = sum(CandidateCount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(PartyName = PresidentialBloc)

seat_results_presidential <- presidential_dapil_stats %>%
  left_join(seat_per_dapil, by = c("ProvinceNorm", "District")) %>%
  rename(Seats = EstimatedSeats) %>%
  group_by(ProvinceNorm, District) %>%
  group_modify(~ allocate_seats(.x)) %>%
  ungroup()

presidential_internal_seat_results <- party_dapil_stats %>%
  add_presidential_coalition(party_col = "PartyName", new_col = "PresidentialBloc") %>%
  mutate(PresidentialBloc = replace_na(PresidentialBloc, "Other")) %>%
  left_join(
    seat_results_presidential %>%
      select(ProvinceNorm, District, PresidentialBloc, BlocSeats = SeatsWon, TotalVotesBloc = TotalVotes),
    by = c("ProvinceNorm", "District", "PresidentialBloc")
  ) %>%
  filter(!is.na(BlocSeats), BlocSeats > 0) %>%
  mutate(Seats = BlocSeats) %>%
  group_by(ProvinceNorm, District, PresidentialBloc) %>%
  group_modify(~ allocate_seats(.x)) %>%
  ungroup() %>%
  select(-Seats)

winners_presidential <- candidate_rankings %>%
  add_presidential_coalition(party_col = "PartyName", new_col = "PresidentialBloc") %>%
  mutate(PresidentialBloc = replace_na(PresidentialBloc, "Other")) %>%
  left_join(
    presidential_internal_seat_results %>%
      select(ProvinceNorm, District, PartyName, PresidentialBloc, SeatsWon, TotalVotes),
    by = c("ProvinceNorm", "District", "PartyName", "PresidentialBloc")
  ) %>%
  mutate(SeatsWon = replace_na(SeatsWon, 0L)) %>%
  filter(ListPosition <= SeatsWon, SeatsWon > 0) %>%
  mutate(
    VoteShare = if_else(TotalVotes > 0, CandidateVote / TotalVotes, NA_real_),
    Province = ProvinceNorm
  ) %>%
  select(Province, District, PresidentialBloc, PartyName, CandidateName, CandidateVote, VoteShare)

winners_by_presidential <- seat_results_presidential %>%
  group_by(PresidentialBloc) %>%
  summarise(EstimatedSeats = sum(SeatsWon, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(EstimatedSeats))

total_seats_presidential <- sum(winners_by_presidential$EstimatedSeats, na.rm = TRUE)

bloc_party_breakdown <- presidential_internal_seat_results %>%
  group_by(PresidentialBloc, PartyName) %>%
  summarise(EstimatedSeats = sum(SeatsWon, na.rm = TRUE), .groups = "drop") %>%
  arrange(PresidentialBloc, desc(EstimatedSeats))

party_seat_comparison <- winners_by_party %>%
  select(PartyName, original_seats = EstimatedSeats) %>%
  full_join(
    presidential_internal_seat_results %>%
      group_by(PartyName) %>%
      summarise(coalition_seats = sum(SeatsWon, na.rm = TRUE), .groups = "drop"),
    by = "PartyName"
  ) %>%
  mutate(across(c(original_seats, coalition_seats), replace_na, 0L)) %>%
  mutate(delta = coalition_seats - original_seats) %>%
  arrange(desc(coalition_seats))
```

## Estimated Winner Snapshot

```{r}
winners %>%
  slice_head(n = 20) %>%
  mutate(
    CandidateVote = comma(CandidateVote),
    VoteShareParty = percent(VoteShareParty, accuracy = 0.1)
  ) %>%
  relocate(PartyName, CandidateName, Province, District)
```

## Seats by Party

Total seats distributed: `r total_seats_original`.

```{r}
winners_by_party %>%
  mutate(PartyName = forcats::fct_reorder(PartyName, EstimatedSeats, .desc = TRUE)) %>%
  ggplot(aes(x = PartyName, y = EstimatedSeats, fill = PartyName)) +
  geom_col() +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Estimated Seat Winners by Party",
    x = "Party",
    y = "Estimated seats"
  ) +
  theme_minimal(base_size = 12) +
  guides(fill = "none")
```

## Party Totals

```{r}
winners_by_party
```

## Coalition Scenario: Sub-4% Parties Unite

The parties below fall under the 4% national vote-share bar and are treated as one bloc in this illustrative rerun of the Sainte-Laguë allocation.

```{r}
party_vote_share %>%
  filter(PartyName %in% coalition_parties) %>%
  arrange(desc(Share)) %>%
  mutate(
    Share = percent(Share, accuracy = 0.1),
    TotalVotesNational = comma(TotalVotesNational)
  )
```

```{r}
winners_by_party_coalition %>%
  mutate(
    PartyName = forcats::fct_reorder(PartyName, EstimatedSeats, .desc = TRUE),
    grouping = if_else(PartyName == coalition_label, coalition_label, "Parliamentary Parties")
  ) %>%
  ggplot(aes(x = PartyName, y = EstimatedSeats, fill = grouping)) +
  geom_col() +
  scale_fill_manual(
    values = c("Parliamentary Parties" = "#5dade2", coalition_label = "#d98880"),
    breaks = c("Parliamentary Parties", coalition_label)
  ) +
  labs(
    title = "Seat Allocation with a Unified Non-Parliamentary Bloc",
    x = "Party",
    y = "Estimated seats",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

Total seats in this scenario: `r total_seats_coalition`.

```{r}
coalition_breakdown %>%
  mutate(OriginalParty = forcats::fct_reorder(OriginalParty, EstimatedSeats, .desc = TRUE)) %>%
  ggplot(aes(x = OriginalParty, y = EstimatedSeats, fill = OriginalParty)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Seats Won Inside the Coalition",
    x = "Party",
    y = "Estimated seats"
  ) +
  theme_minimal(base_size = 12) +
  guides(fill = "none")
```

```{r}
coalition_breakdown
```

## Presidential Coalition Scenario

Parties are grouped by the presidential ticket they backed during the 2024 race.

```{r}
tibble(Party = names(presidential_map), PresidentialBloc = unname(presidential_map)) %>%
  arrange(PresidentialBloc, Party)
```

Total seats across blocs: `r total_seats_presidential`.

```{r}
winners_by_presidential %>%
  mutate(PresidentialBloc = forcats::fct_reorder(PresidentialBloc, EstimatedSeats, .desc = TRUE)) %>%
  ggplot(aes(x = PresidentialBloc, y = EstimatedSeats, fill = PresidentialBloc)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Anies" = "#1b998b",
    "Prabowo" = "#ff7f0e",
    "Ganjar" = "#c93a3a",
    "Independent" = "#7d3c98",
    "Other" = "#95a5a6"
  )) +
  labs(
    title = "Seat Allocation by Presidential Coalition",
    x = "Presidential bloc",
    y = "Estimated seats"
  ) +
  theme_minimal(base_size = 12) +
  guides(fill = "none")
```

```{r}
bloc_party_breakdown %>%
  group_by(PresidentialBloc) %>%
  mutate(PartyFacet = forcats::fct_reorder(PartyName, EstimatedSeats, .desc = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = PartyFacet, y = EstimatedSeats, fill = PartyName)) +
  geom_col() +
  facet_wrap(~ PresidentialBloc, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Party Contributions Inside Each Bloc",
    x = "Party",
    y = "Estimated seats"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none")
```

```{r}
winners_presidential %>%
  group_by(PresidentialBloc) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(
    CandidateVote = comma(CandidateVote),
    VoteShare = percent(VoteShare, accuracy = 0.1)
  ) %>%
  arrange(PresidentialBloc, desc(as.numeric(gsub(",", "", CandidateVote))))
```

### Party Seat Counts Under Presidential Blocs

```{r}
party_seat_comparison %>%
  mutate(
    original_seats = comma(original_seats),
    coalition_seats = comma(coalition_seats),
    delta = scales::number(delta, suffix = " seats", big.mark = ",", style_positive = "plus")
  ) %>%
  rename(
    `Party` = PartyName,
    `Original Seats` = original_seats,
    `Coalition Seats` = coalition_seats,
    `Change` = delta
  )
```
